Neural Connectome of the Ctenophore Statocyst

### 2025-07-08 manuscript/Manuscript.qmd
edit main text (Results, Discussion and Materials and Methods)

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To obtain high-quality ultrastructural preservation, we used high pressure freezing followed by freeze substitution and resin embedding (Epon). The specimen was sectioned from the aboral tip and we collected approximately 1,000 ultra-thin (50 nm) serial sections as ribbons. The sections were imaged in a Zeiss Gemini 500 scanning electron microscope at 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016], resulting in a reconstruction of 972 cells that had intact nuclei and cell bodies, 898 of which formed the aboral organ (AO). Most of the 898 AO cells, including the balancers and the aboral nerve net neurons are fully contained within the volume, due to the compact and mostly self-contained organisation of the aboral organ. We classified these cells into cell types (Figure 1--figure supplement 1, Figure 1--figure supplement 2).

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies (Video 2 and not shown). For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), wherein separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. In such cases, branch nodes were placed near the closest existing node and annotated accordingly (CATMAID only supports skeleton trees). The entire skeletonized volume was composed of 134,591 nodes. 73 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

@@

Immunofuorescent stainings in *Pleurobrachia pileus* revealed a nerve net at the base of the aboral organ that seemed continuous with the polygonal nerve net of the surrounding epithelium [@Jager2011]. Ultrastructural studies revealed synapses near the floor of the aboral organ—a region where epithelial cells are densely packed and form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a]. In our dataset, we identified 428 synapses based on the previously described pre-synaptic triad morphology consisting of a single layer of vesicles, a cisterna of smooth endoplasmic reticulum and an associated mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane (light blue arrows in CATMAID) (Figure 2--figure supplement 1A). We could not detect specialized post-synaptic structures, in agreement with previous studies [@Hernandez_Nicaise1973a]. Synapses were identified as either monoadic or polyadic, with one pre-synaptic neuron forming connections with one or multiple post-synaptic cells (Figure 2--figure supplement 1B).

We reconstructed three aboral nerve-net (ANN) neurons, each with a syncytial morphology (Figure 2A and B). Each ANN had multiple nuclei (ranging from two to five per neuron), with anastomosing membranes (Figure 2--figure supplement 1C). ANN neurons are distinct from the syncytial subepithelial nerve net (SNN) neurons with blebbed morphology previously identified by vEM in the body wall of *Mnemiopsis* [@Burkhardt2023]. ANN neurons also lack dense core vesicles, structures that are abundant throughout the cell bodies and neurites of SNN neurons [@Burkhardt2023]. In our dataset, we also reconstructed 10 SNN neurons. These cells showed a blebbed morphology, contained several dense core vesicles and were located at the periphery of the AO (Figure 2--figure supplement 1). Our results reveal a second type of syncytial neuron in ctenophores (Figure 2--figure supplement 1 C-G).

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on their distinct size and position, we classified the ANNs into two types. The first type is a single large neuron (ANN_Q1-4) that spans all four quadrants (Q1-Q4) and has four (possibly five) nuclei. This neuron has 150 presynaptic sites (Figure 2C-F). The second type consists of two neurons (ANN_Q1Q2 and ANN_Q3Q4), each spanning two quadrants and containing two nuclei. ANN_Q1Q2 had 119 presynapses, whereas ANN_Q3Q4 had 68 presynaptic structures (Figure 2C-F). SNN neurons had no or only few synapses.

@@

Our synapse annotation revealed several types of synaptic connections within the gravisensory organ. We identified synapses between different ANN neurons and synapses that ANN_Q1-4 forms on itself (autapses). The ANN neurons also form synapses on the gravity-sensing balancer cells and the bridge cells. Bridge cells, but not balancer cells, also synapse on the ANN.

@@

ANN_Q1-4 formed synapses on balancer cells in all four quadrants (on 7 cells in Q1, 11 cells in Q2, 6 cells in Q3 and 10 cells in Q4) while ANN_Q1Q2 and ANN_Q3Q4 synapsed to balancer cells in their respective quadrants (ANN_Q1Q2 on 6/37 cells in Q1 and 8/32 cells in Q2; ANN_Q3Q4 on 1/32 cells in Q3 and 5/28 cells in Q4). Some balancer cells received inputs from both ANN_Q1-4 and either of ANN_Q1Q2 or ANN_Q3Q4. While previous studies suggested the presence of afferent synapses from balancer cells to neurons [@Hernandez_Nicaise1974], our data did not reveal any presynaptic sites in balancer cells (Figure 3 -- figure supplement 1).

@@

@@

To analyze the synaptic connectivity graph of the balancer organ and the ANN, we grouped cells of the same type and in the same region into a single node, summed the number of synapses, and laid out the network to reflect the anatomy of the four quadrants (Figure 3H). The network is characterised by feedback connections among ANN neurons and between ANN and bridge cells with a clear rotational (or mirror) symmetry. Q1Q2 and Q3Q4 also form separate local circuits centered around their regional ANN neurons while ANN_Q1-4 mediates global connectivity across the entire balancer organ. Notably, there were no synapses from the mechanosensory balancer cells to the ANN or from ANN to motor cells (e.g. ciliary groove), contrary to our initial expectation of a sensory-motor or input-output model of balancer function.

@@

To investigate the regulation of balancer function, we carried out a high-speed video-microscope analysis of balancer cilia. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015]. To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).

@@

During the two-minute recordings, balancer cilia could beat fast, slow, or exhibit abrupt stops (arrest) and start moving again (re-beat). Occasionally, large body-contractions moved the entire cydippid out of frame, and data from these episods were excluded. We focused on three metrics for inter-quadrant comparison: (1) the timing of ciliary arrest, (2) the timing of re-beat after an arrested phase, and (3) ciliary beat frequency.

@@

As shown in Figure 4E (and Figure 4--figure supplement 1), beat frequencies of left and right balancers fluctuated substantially in both orientations. However, while left-right activity was largely synchronized in sagittal pairs (Figure 4E, left), tentacular pairs (Figure 4E, right) showed strikingly periodic alternations between left and right balancer activity. This pattern suggested an offset in timing during arrest or re-beat events between tentacular balancers.

To quantify this, we calculated a rolling Pearson correlation between left and right balancer activity in each plane. The resulting mean correlations were significantly higher in the tentacular plane compared to the sagittal plane (Figure 4F; t-test, p = 0.014). This result indicates that tentacular balancers beat in a more coordinated, mirror-symmetric manner, whereas sagittal balancers exhibit greater variability between sides.

@@

Notably, bridge neurons associated with each quadrant region (Q1Q2 or Q3Q4) extend their processes across the midline, forming connections that span between opposite quadrant domains when viewed along the tentacular plane. This morphology suggests a potential role for bridge neurons in modulating inter-quadrant coordination. By linking ANN activity across orthogonal planes, these neurons may contribute to the dynamic regulation of CBF patterns across the whole system.

![**Figure 4. Analysis of ciliary beating, arrests and re-beat across balancers in the four quadrants**\
(A) Schematic diagram of the differential interference contrast (DIC) microscopy setup used to image the movement of balancer cilia. The microscope was tilted 90 degrees so that the stage was positioned vertically. We used a 40x objective lens and a monochrome CMOS camera sensitive to near-infrared (NIR) light that was synchronized with an 850 nm strobe light source.\
(B) Representative DIC images of the aboral organ viewed along the sagittal (left) and tentacular (right) planes. Arrowheads indicate the balancer cilia selected for kymograph-based analysis in each orientation. Scale bar: 25 µm.\
(C) Representative kymographs showing arrest and re-beat events of balancer cilia. Left and right balancer cilia were simultaneously recorded in either the sagittal (left) or tentacular plane (right). Arrowheads mark the onset of arrest (open) and re-beat (filled); horizontal arrows represent 1-second intervals.\
(D) Boxplots showing time differences in the onset of arrest (white boxes) and re-beat (grey boxes) between left and right balancer cilia. Colors of the box outlines indicate the imaging plane: sagittal (blue) or tentacular (orange). Each dot represents a single larva.\
(E) Bar plots showing ciliary beat frequency (CBF) dynamics of left and right balancers (top and bottom, respectively) recorded from the sagittal (left) or the tentacular (right) planes. Positive and negative values represent opposite sides within each plane.\
(F) Mean rolling Pearson correlation of left–right balancer CBF calculated with a 20-frame window. Each dot represents an individual larva. Balancers on the tentacular plane showed significantly higher correlation than those on the sagittal plane (t-test, p = 0.014).](figures/Figure4.png)

@@

In this study, we generated the first high-resolution connectome of the ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of a five-day-old *Mnemiopsis leidyi* cydippid by volumetric electron microscopy. Among the reconstructed cells, we identified a previously undescribed type of syncytial neuron---here referred to as aboral nerve net (ANNs)---with multiple nuclei. The ANN has a morphology distinct from that of the subepithelial nerve net (SNN) reported for one-day-old cydippids [@Burkhardt2023; @Sachkova2021]. Instead of forming discrete nodal swellings or a “beads-on-a-string” organization typical for the SNN, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion. The presence of two SNN-type neurons more laterally in the aboral organ in our dataset (Figure 2--figure supplement1) further underline the distinct identity of ANNs.

@@

Our findings reveal a circuit composed of ANN neurons and bridge cells that coordinates ciliary beating rhythms in space and time without sensory input, and features a recursive architecture in which ciliary outputs may feed back to the circuit via bridge cells. Such a configuration suggests that even with a limited number of cells, the system is capable of flexible and robust rhythm control.

As Jagar et al. (2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011]. The network we describe here orchestrates output through dynamic, self-regulating coordination. Future work could uncover this dynamics, visualized through live imaging of the entire cell ensemble, to offer a window into the core functions of a unique nervous system.

@@

Larvae of *Mnemiopsis leidyi* (five days old) were cryofixed using a high-pressure freezing apparatus (HPM Live μ, CryoCapCell, https://www.cryocapcell.com/technologies/high-pressure-freezer) and immediately transferred to liquid nitrogen for storage. The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica). Cryo-substitution was carried out by gradually raising the temperature from -90°C to -70°C over 4 hours, then returning to -20°C over 2 hours, and finally to room temperature over 2 hours. The samples were then embedded in epoxy resin (EMbed 812, Electron Microscopy Sciences). Serial ultrathin sections of 50 nm thickness were prepared using a Leica UC7 ultramicrotome and a 45º DiATOME diamond knife. To enhance section adhesion and improve hydrophilicity, a conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.), rendering the surface negatively charged. Section ribbons were collected on prepared glass slides, gently dried to ensure proper stretching, and firmly adhered to the glass surface. The sections were stained with UranyLess and lead citrate (Reynolds) using airless staining procedures. The glass slides were mounted on a STEM-specific stage (Zeiss) using Copper Foil EMI Shielding Tape (3M). Imaging was performed using a Gemini SEM 500 (Zeiss), equipped with SmartSEM and Atlas 5 imaging software (Zeiss), using the Inlens detector. A total of 620 serial sections, each 50 nm thick, were analyzed. The imaging resolution was 2.8 nm/pixel, and the acceleration voltage was set to 1.5 kV. The imaging time for each tile was 54 minutes and 21 seconds, with a pixel size of 2.8 nm, a tile size of 91.8 × 91.8 µm (32768 × 32768 pixels), and a dwell time of 3.0 µs.

@@

We assigned each cell a specific cell-type name based on its category (balancer, bridge, large granular, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, ANN) resulting in a total of 12 cell types. Some cells only had more general features and were classified based on the presence/absence and number of cilia into four ciliated cell types:(biciliated (biC), monociliated (monoC), multiciliated (multiC), non-ciliated (non-C)). Additionally, we appended the quadrant number to the cell type name after an underscore ("\_") to indicate the cell-body’s location. For example, cells located in the first quadrant are named with "\_Q1." If a cell is situated between the first and second quadrants, we appended "Q1Q2" to the cell-type name.

@@

For ciliary imaging, we used the cydippid stage of *M. leidyi* at five days post-fertilization. A coverslip with a thin layer of Vaseline applied to its two edges was gently placed over a glass slide. Filtered natural seawater and the cydippids were introduced beneath the coverslip. The orientation of the larvae was adjusted by carefully sliding the coverslip, and slight pressure was then applied to immobilize the larvae.

Balancer cilia movement was imaged using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a high-sensitivity monochrome CMOS camera optimized for the near-infrared range (UI-3360CP-NIR-GL Rev.2, iDS), and a 40x glycerine-immersion objective lens (LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27). To stabilize the statolith, the microscope was tilted by 90° to achieve vertical stage alignment. Illumination was provided by a custom-made NIR LED strobe system (850 nm) synchronized with the camera’s exposure signals. Image acquisition was controlled using BohNavi software. Recordings were acquired at 640 × 480 pixels, 100 frames per second (fps), for a duration of 2 minutes, using 0.05 ms pulse illumination from the LED.

For beat frequency analysis, kymographs were generated using the Multi Kymograph plugin in Fiji. One-second segments were manually extracted from the kymographs, and ciliary beat frequencies were counted visually for each ROI. These manually determined frequencies were then used for subsequent computational analysis.

To process the beat frequency time series data, left and right cilia traces were extracted, filtered to remove missing or non-finite values, and assigned frame indices. Data were saved as RDS objects for downstream analysis. Mean values were plotted as bar graphs with the left and right cilia shown in positive and negative directions, respectively.

@@

This work was supported by the Japan Society for the Promotion of Science (JSPS) Overseas Research Fellowships, the Grass Foundation, the Kavli Foundation, and the Marine Biological Laboratory (MBL). This project also received funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement No. 101020792). We thank Dr. Réza Shahidi for sectioning and imaging, Paulina Cherek for high-pressure freezing and sample preparation, and Iva Verbanac for help with tracing. We are also grateful to Dr. Chris Bjornsson, the MBL Central Microscopy Facility, Dr. Shoji A. Baba, and Dr. Kogiku Shiba (Shimoda Marine Research Center) for their assistance with microscopy.

>>>>

For vEM analysis, we used a five-day-old cydippid larva of *Mnemiopsis leidyi*. To ensure optimal ultrastructural preservation, the specimen was prepared by high-pressure freezing, followed by freeze substitution and Epon embedding. From the aboral tip, we collected \~1,000 ultra-thin (50 nm) serial sections as ribbons, targeting only the region containing the aboral organ. These were imaged using a Zeiss Gemini 500 SEM at a resolution of 2.0 nm/pixel (xy). We imaged only the region containing the aboral organ. Subsequently, images from 620 layers were stitched and aligned in TrakEM2 [@cardona2012]. The final vEM dataset encompassed a volume of 60 μm × 40 μm × 30 μm. We skeletonised and annotated all cells in this dataset in CATMAID [@Saalfeld2009; @Schneider-Mizell2016]. We reconstructed 972 cells with intact nuclei and somata. Among these, 898 were located within the aboral organ (AO), including balancers and the aboral nerve net neurons. Due to the compact and mostly self-contained organization of the AO, these cells were largely contained within the imaged volume. We classified these cells into cell types (Figure 1--figure supplement 1, Figure 1--figure supplement 2).

In cells containing cilia, we also traced cilia along their length and annotated all basal bodies (Video 2 and not shown). For neuronal skeletonization, nodes were placed to interconnect the profiles of the same neuron’s processes across layers, extending the skeleton until all branches were fully traced. All nodes relevant to synaptic sites were tagged, and skeletons were named and assigned multi-level annotations. As described later, some neurons formed loop-like structures (anastomosed neurons), in which separated branches often rejoined either the main branch or smaller branches [@Burkhardt2023]. Since CATMAID only supports skeleton trees, branch nodes in such cases were placed near the closest existing node and annotated accordingly. The entire skeletonized volume was composed of 134,591 nodes. 73 fragments could not be attached to somata-associated skeletons. Most of these fragments represent short skeletal branches that could not be traced beyond gaps or low-quality layers.

@@

Previous immunofluorescent staining in *Pleurobrachia pileus* suggested the presence of a nerve net at the base of the aboral organ, possibly continuous with the surrounding epithelial nerve net [@Jager2011]. Electron microscopy studies later revealed synapses near the floor of the aboral organ, a region where densely packed epithelial cells form a floor-like structure [@Horridge_1964; @Hernandez_Nicaise1973a].

In our volume EM dataset, we identified 428 synapses based on the canonical presynaptic triad morphology: a cluster of synaptic vesicles, a smooth endoplasmic reticulum cisterna, and a mitochondrion. At synaptic sites, we marked mitochondria as synaptic nodes (orange in CATMAID) and connected this node to the nearest node in postsynaptic cells across the regions where synaptic vesicles aligned with the presynaptic cell's membrane, as indicated by light blue arrows in CATMAID (Figure 2--figure supplement 1A). No specialized postsynaptic densities were observed, consistent with prior reports [@Hernandez_Nicaise1973a]. Synapses were either monoadic (one postsynaptic target) or polyadic (multiple postsynaptic targets) (Figure 2--figure supplement 1B).

We reconstructed three aboral nerve-net (ANN) neurons, each exhibiting a syncytial morphology characterized by anastomosing membranes and multiple nuclei (ranging from two to five) (Figure 2A and B, Figure 2--figure supplement 1C). These ANN neurons are morphologically distinct from the previously described syncytial subepithelial nerve net (SNN) neurons in the body wall of *Mnemiopsis*, which show a blebbed morphology and contain abundant dense core vesicles [@Burkhardt2023]. Our results identify a second type of syncytial neuron in *Mnemiopsis*. In addition to the previously characterized SNN neurons in the body wall, we observed ANN neurons within the aboral organ. Unlike SNNs, which exhibit a blebbed morphology and contain numerous dense core vesicles, ANN neurons are smooth-surfaced and lack dense vesicles. ANN neurons were located centrally within the aboral organ, while SNN neurons were reconstructed near its periphery (Figure 2--figure supplement 1C–G).

Furthermore, ANNs are also different from other neurons previously reported by EM, including mesogleal neurons and ciliated sensory cells that synapse on the SNN (types 1-4)[@Burkhardt2023]. Based on soma size and spatial distribution, we classified ANN neurons into two types. The first is a single large neuron, ANN_Q1-4, spanning all four quadrants and containing four (possibly five) nuclei. This neuron formed 150 presynaptic sites. The second type includes two smaller neurons, ANN_Q1Q2 and ANN_Q3Q4, each confined to two adjacent quadrants and containing two nuclei. ANN_Q1Q2 had 119 presynapses, while ANN_Q3Q4 had 68 (Figure 2C–F). In contrast, SNN neurons exhibited either no synaptic contacts or only a few.

@@

Synapse annotation revealed multiple types of connections involving ANN neurons within the gravisensory organ, including reciprocal ANN–bridge cell connections, ANN autapses, and outputs to balancer cells. We identified synapses between different ANN neurons and synapses that ANN_Q1-4 forms on itself (autapses). The ANN neurons also form synapses on the gravity-sensing balancer cells and the bridge cells. Bridge cells, but not balancer cells, also synapse on the ANN.

@@

ANN_Q1-4 formed synapses on balancer cells in all four quadrants (on 7 cells in Q1, 11 cells in Q2, 6 cells in Q3 and 10 cells in Q4) while ANN_Q1Q2 and ANN_Q3Q4 synapsed to balancer cells in their respective quadrants (ANN_Q1Q2 on 6/37 cells in Q1 and 8/32 cells in Q2; ANN_Q3Q4 on 1/32 cells in Q3 and 5/28 cells in Q4). Some balancer cells received inputs from both ANN_Q1-4 and either of ANN_Q1Q2 or ANN_Q3Q4. Notably, we found no presynaptic sites in balancer cells (Figure 3 -- figure supplement 1), contradicting earlier suggestions of afferent signaling from balancer cells to neurons [@Hernandez_Nicaise1974].

@@

In contrast to balancer cells, bridge cells participated in reciprocal synaptic interactions with ANN neurons, forming both pre- and postsynaptic connections.

@@

To examine the overall synaptic organization of the balancer organ, we grouped cells of the same type and region into network nodes, summed the synaptic counts, and mapped the network onto the four-quadrant anatomy (Figure 3H). The resulting connectivity graph revealed feedback loops among ANN neurons and between ANN and bridge cells, displaying rotational symmetry. Local sub-circuits centered around ANN_Q1Q2 and ANN_Q3Q4 were evident, while ANN_Q1-4 formed global connections across quadrants.

Surprisingly, and contrary to the expected sensory–motor or input–output model of balancer function, we found no synapses from the mechanosensory balancer cells to ANN neurons, nor from ANN neurons to known motor effectors such as ciliated groove cells.

@@

To investigate whether balancer cilia are coordinated by neural inputs, we performed high-speed video microscopy to analyze their activity patterns across body axes. Previous studies [@Tamm1980; @Tamm1982; @Lowe1997] have established that balancer cilia function as mechanoreceptors, with their beating frequency modulated by inclination. Tamm also suggested that differences in statolith morphology and the shape of balancer cilia between the tentacular and sagittal planes could lead to different forces exerted on cilia by the statolith [@Tamm_2014c; @tamm2015]. To further explore this, we used a tilted microscope with a vertical stage where we mounted immobilised cydippid larvae with their aboral-oral axis aligned in different orientations relative to the gravity vector. We then compared larvae oriented in different angles and either with their sagittal or tentacular plane parallel to the sample stage (Figure 4A).

@@

During the two-minute recordings, balancer cilia could beat fast, slow, or exhibit abrupt stops of beating ("arrest") followed by re-initiation ("re-beat"). Occasionally, large body-contractions moved the entire cydippid out of frame, and data from these episods were excluded. We focused on three metrics for inter-quadrant comparison: (1) the timing of ciliary arrest, (2) the timing of re-beat after an arrested phase, and (3) ciliary beat frequency.

@@

As shown in Figure 4E (and Figure 4--figure supplement 1), beat frequencies of left and right balancers fluctuated substantially in both orientations. However, while left–right activity was largely synchronized in sagittal pairs (Figure 4E, left), tentacular pairs (Figure 4E, right) displayed strikingly periodic alternations between left and right balancer activity. This pattern suggests a consistent temporal offset between the two sides, possibly arising from differences in the timing of arrest or re-beat events.

To quantify this, we calculated a rolling Pearson correlation between left and right balancer activity in each plane. The resulting mean correlations were significantly higher in the tentacular plane compared to the sagittal plane (Figure 4F; t-test, *p* = 0.014). This result indicates that tentacular balancers beat in a more coordinated, mirror-symmetric manner, whereas sagittal balancers exhibit greater variability between sides.

@@

Notably, bridge neurons associated with each quadrant region (Q1Q2 or Q3Q4) extend their processes across the midline, forming connections that span between opposite quadrant domains when viewed along the tentacular plane. This morphology suggests a potential role for bridge neurons in modulating inter-quadrant coordination. By linking ANN activity across orthogonal planes, these neurons may contribute to the dynamic regulation of CBF patterns by facilitating information flow across orthogonal planes, thus supporting coordinated activity throughout the aboral organ.

![**Figure 4. Analysis of ciliary beating, arrests and re-beat across balancers in the four quadrants**\
(A) Schematic diagram of the differential interference contrast (DIC) microscopy setup used to image the movement of balancer cilia. The microscope was tilted 90 degrees so that the stage was positioned vertically. We used a 40x objective lens and a monochrome CMOS camera sensitive to near-infrared (NIR) light that was synchronized with an 850 nm strobe light source.\
(B) Representative DIC images of the aboral organ viewed along the sagittal (left) and tentacular (right) planes. Arrowheads indicate the balancer cilia selected for kymograph-based analysis in each orientation. Scale bar: 25 µm.\
(C) Representative kymographs showing arrest and re-beat events of balancer cilia. Left and right balancer cilia were simultaneously recorded in either the sagittal (left) or tentacular plane (right). Arrowheads mark the onset of arrest (open) and re-beat (filled); horizontal arrows represent 1-second intervals.\
(D) Boxplots showing time differences in the onset of arrest (white boxes) and re-beat (grey boxes) between left and right balancer cilia. Colors of the box outlines indicate the imaging plane: sagittal (blue) or tentacular (orange). Each dot represents a single larva.\
(E) Bar plots showing ciliary beat frequency (CBF) dynamics of left and right balancers (top and bottom, respectively) recorded from the sagittal (left) or the tentacular (right) planes. Positive and negative values represent opposite sides within each plane.\
(F) Mean rolling Pearson correlation of left–right balancer CBF calculated with a 20-frame window. Each dot represents an individual larva. Balancers on the tentacular plane showed significantly higher correlation than those on the sagittal plane (t-test, *p* = 0.014).](figures/Figure4.png)

@@

In this study, we generated the first high-resolution connectome of the ctenophore nervous system by reconstructing approximately 900 cells from the aboral organ of a five-day-old *Mnemiopsis leidyi* cydippid by volumetric electron microscopy. Among the reconstructed cells, we identified a previously undescribed type of syncytial neuron---here referred to as aboral nerve net (ANNs)---with multiple nuclei. The ANN has a morphology distinct from that of the subepithelial nerve net (SNN) reported for one-day-old cydippids [@Burkhardt2023; @Sachkova2021]. A recent study also identified a distinct subtype of SNN neurons in the aboral region, termed “condensed” SNN, which display a more compact morphology compared to body wall SNNs [@Ferraioli2025]. However, the ANN neurons described here are morphologically and anatomically distinct from both the body wall SNN and the condensed SNN. Instead of forming discrete nodal swellings or a “beads-on-a-string” organization typical for the SNN, ANNs extend through intercellular spaces, weaving between neighboring cells in a smooth, interdigitating fashion. The presence of two SNN-type neurons more laterally in the aboral organ in our dataset (Figure 2--figure supplement1) further underline the distinct identity of ANNs.

@@

Our findings reveal a circuit composed of ANN neurons and bridge cells that coordinates ciliary beating rhythms in space and time without direct sensory input, and features a recursive architecture in which ciliary outputs may feed back to the circuit via bridge cells. Such a configuration suggests that even with a limited number of cells, the system is capable of flexible and robust rhythm control.

As Jagar et al. (2011) once described, this organ evokes the metaphor of a “ciliary brain” [@Jager2011]. The network we describe here orchestrates output through dynamic, self-regulating coordination. Future work could uncover these dynamics, visualized through live imaging of the entire cell ensemble, to offer a window into the core functions of a unique nervous system.

@@

Larvae of *Mnemiopsis leidyi* (five days old) were cryofixed using a high-pressure freezing apparatus (HPM Live μ, CryoCapCell, https://www.cryocapcell.com/technologies/high-pressure-freezer) and immediately transferred to liquid nitrogen for storage. The frozen samples were processed in a substitution medium containing 2% (w/v) osmium tetroxide and 0.5% uranyl acetate in acetone, using a cryo-substitution device (EM AFS-2, Leica).

Cryo-substitution was carried out by gradually raising the temperature from -90°C to -70°C over 4 hours, then returning to -20°C over 2 hours, and finally to room temperature over 2 hours. The samples were then embedded in epoxy resin (EMbed 812, Electron Microscopy Sciences).

Serial ultrathin sections of 50 nm thickness were prepared using a Leica UC7 ultramicrotome and a 45º DiATOME diamond knife. A conductive indium tin oxide-coated glass slide (ITO Glass, UQG Optics) was treated with air glow discharge using the PELCO easiGlow system (Ted Pella, Inc.) to enhance section adhesion and improve hydrophilicity. This process rendered the glass surface negatively charged.

Section ribbons were collected on prepared glass slides, gently dried to ensure proper stretching, and firmly adhered to the glass surface. The sections were stained with UranyLess and lead citrate (Reynolds) using airless staining procedures. The glass slides were mounted on a STEM-specific stage (Zeiss) using Copper Foil EMI Shielding Tape (3M).

Imaging was performed using a Gemini SEM 500 (Zeiss), equipped with SmartSEM and Atlas 5 imaging software (Zeiss), using the Inlens detector. A total of 620 serial sections, each 50 nm thick, were analyzed. The imaging resolution was 2.8 nm/pixel, and the acceleration voltage was set to 1.5 kV. The imaging time for each tile was 54 minutes and 21 seconds, with a pixel size of 2.8 nm, a tile size of 91.8 × 91.8 µm (32768 × 32768 pixels), and a dwell time of 3.0 µs.

@@

We assigned each cell a specific cell-type name based on its morphological or functional category (e.g., balancer, bridge, large granular, ciliated grooves, dense vesicle cell, dome, epithelial floor (ef), intracellular multiciliated cells (imc), lamellate bodies (lb), lithocyte, plumose cell, and ANN), resulting in a total of 12 primary cell types.

In addition, some cells lacked distinctive features and were instead categorized into one of four general ciliated types, based on the presence and number of cilia: biciliated (biC), monociliated (monoC), multiciliated (multiC), or non-ciliated (non-C).

To indicate spatial positioning, we appended the quadrant number to the cell-type name using an underscore. For instance, cells located in the first quadrant were labeled "\_Q1," while those situated between two quadrants were labeled with both (e.g., "Q1Q2").

Cells of the same type and in the same region were further distinguished by serial numbering, ensuring each cell has a unique identifier.

@@

For ciliary imaging, we used the cydippid stage of *M. leidyi* at five days post-fertilization. A coverslip with a thin layer of Vaseline applied to its two edges was gently placed over a glass slide. Filtered natural seawater and cydippids were placed beneath the coverslip. The orientation of the larvae was adjusted by carefully sliding the coverslip, and the larvae were gently immobilized by applying slight pressure.

To ensure consistency in gravitational input, we included only larvae whose aboral–oral axis was oriented nearly perpendicular to the gravitational vector (tilted by no more than 20°). This minimized the potential influence of larval tilt on left–right differences in balancer cilia activity, as all balancers received comparable stimulation from the statolith.

Balancer cilia movement was imaged using a differential interference contrast (DIC) microscope (Zeiss Axio Imager.M2) equipped with a high-sensitivity monochrome CMOS camera optimized for the near-infrared range (UI-3360CP-NIR-GL Rev.2, iDS), and a 40x glycerine-immersion objective lens (LD LCI Plan-Apochromat 40x/1.2 Imm Corr DIC M27). To ensure consistent statolith positioning during imaging, the microscope was tilted by 90° to achieve vertical stage alignment. Illumination was provided by a custom-made NIR LED strobe system (850 nm) synchronized with the camera’s exposure signals. Image acquisition was controlled using BohNavi software. Recordings were acquired at 640 × 480 pixels, 100 frames per second (fps), for a duration of 2 minutes, using 0.05 ms pulse illumination from the LED.

For beat frequency analysis, kymographs were generated using the Multi Kymograph plugin in Fiji. One-second segments were manually extracted from the kymographs, and ciliary beat frequencies were visually counted for each manually defined region of interest (ROI). These manually determined frequencies were then used for subsequent computational analysis.

Traces of left and right cilia were extracted, cleaned to remove missing or non-finite values, and indexed by frame. Data were saved as RDS objects for downstream analysis. Mean values were plotted as bar graphs with the left and right cilia shown in positive and negative directions, respectively.

@@

This work was supported by the Japan Society for the Promotion of Science (JSPS) Overseas Research Fellowships, the Grass Foundation, the Kavli Foundation, and the Marine Biological Laboratory (MBL). This project also received funding from the European Research Council (ERC) under the European Union’s Horizon 2020 research and innovation programme (grant agreement No. 101020792). We thank Dr. Réza Shahidi for sectioning and imaging, Paulina Cherek for high-pressure freezing and sample preparation, and Iva Verbanac for help with tracing. We are also grateful to Dr. Chris Bjornsson, the MBL Central Microscopy Facility, Dr. Shoji A. Baba, and Dr. Kogiku Shiba for their assistance with microscopy. We also thank Dr. Sidney L. Tamm and Dr. Steven H. D. Haddock for their valuable feedback on the manuscript.
